BHPWAVE: An adiabatic gravitational waveform model for compact objects undergoing quasi-circular inspirals into rotating massive black holes

### 2023-10-23 ms.tex
Updates from Overleaf

@@

% \begin{figure*}[!htp]
%     \centering
%     \includegraphics[width=0.98\linewidth]{figures/phase_error.pdf}
%     \caption{Different rescalings and reparametrizations of the flux to improve interpolation accuracy of the numerical data.}
%     \label{fig:fisherCompare}
% \end{figure*}

In Figure \ref{fig:fisherCompare} we plot the resulting statistical uncertainties and systematic biases of $(\Delta\log(M/M_\odot), \Delta\log(\mu/M_\odot), \Delta a, \Delta p_0)$ as a function of $a$ for the various sources. Solid lines refer to the statistical uncertainties, while dashed lines refer to the systematic biases. Different colors refer to different combinations of $M$ and $\mu$. We see that the systematic biases always fall under the intrinsic uncertainty estimates. Furthermore, the uncertainties decrease as spin increases, with the biases possessing largely the same behavior. Thus, at least in the region of parameter space considered for this toy analysis, we see that the interpolation errors in the phases and frequencies of our adiabatic trajectories do not significantly bias our waveforms.

@@

In Figure \ref{fig:errorBiases} we plot the dephasing $\Delta \Phi_\mathrm{GW} = 2(\Phi_B - \Phi_E)$ of the two gravitational wave models, the maximum ratio between $\Delta \theta^i_\mathrm{stat}$ and $\Delta \theta^i_\mathrm{bias}$ for the intrinsic parameters $\mathcal{R}_\mathrm{intrinsic}$, the maximum ratio between $\Delta \theta^i_\mathrm{stat}$ and $\Delta \theta^i_\mathrm{bias}$ for the extrinisic parameters $\mathcal{R}_\mathrm{extrinsic}$, along with the mismatch $\mathcal{M}_\mathrm{bf}$ between $h_B(\theta-\Delta \theta_\mathrm{bias})$ and $h_E(\theta)$. Note that we estimate $\Delta \Phi_\mathrm{GW}$ from the dephasing between the $(2,2)$-modes, which is two times the difference between the phase trajectories of the \texttt{bhpwave} and corrupted models, $\Phi_B$ and $ \Phi_D$, respectively. The top panel includes results for the binary with masses $(M, \mu) = (10^6 M_\odot, 10 M_\odot)$, while the bottom panel is for a binary with $(M, \mu) = (5\times 10^6 M_\odot, 10 M_\odot)$.

@@

We presented the theoretical and numerical methods behind \texttt{bhpwave}: a new Python-based adiabatic gravitational waveform generator for binary sources composed of a compact object undergoing a quasi-circular inspiral into a rotating massive black hole. To build this waveform model, we precomputed mass-independent trajectories for systems with initial separations $r_0/M \lesssim 60$ and Kerr spins $|a| \leq 0.9999$. Furthermore, we precomputed all potential waveform harmonic amplitudes for $\ell \leq 20$. By im

>>>>

\begin{figure}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{figures/phase_error.pdf}
    \caption{Different rescalings and reparametrizations of the flux to improve interpolation accuracy of the numerical data.}
    \label{fig:fisherCompare}
\end{figure}

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.98\linewidth]{figures/flux_error.pdf}
    \caption{Different rescalings and reparametrizations of the flux to improve interpolation accuracy of the numerical data.}
    \label{fig:errorBiases}
\end{figure*}

@@

In Figure \ref{fig:fisherCompare} we plot the resulting statistical uncertainties and systematic biases of $(\Delta\log(M/M_\odot), \Delta\log(\mu/M_\odot), \Delta \hat{a}, \Delta p_0)$ as a function of $1-\hat{a}$ for the various sources. Solid lines refer to the statistical uncertainties, while dashed lines refer to the systematic biases. Different colors refer to different combinations of $M$ and $\mu$. We see that the systematic biases always fall under the intrinsic uncertainty estimates. Furthermore, the uncertainties decrease as spin increases, with the biases possessing largely the same behavior. Thus, at least in the region of parameter space considered for this toy analysis, we see that the interpolation errors in the phases and frequencies of our adiabatic trajectories do not significantly bias our waveforms.

@@

In Figure \ref{fig:errorBiases} we plot the dephasing $\Delta \Phi_\mathrm{GW} = 2(\Phi_B - \Phi_E)$ of the two gravitational wave models, the maximum ratio between $\Delta \theta^i_\mathrm{stat}$ and $\Delta \theta^i_\mathrm{bias}$ for the intrinsic parameters $\mathcal{R}_\mathrm{intrinsic}$, the maximum ratio between $\Delta \theta^i_\mathrm{stat}$ and $\Delta \theta^i_\mathrm{bias}$ for the extrinisic parameters $\mathcal{R}_\mathrm{extrinsic}$, along with the mismatch $\mathcal{M}_\mathrm{bf}$ between $h_B(\theta-\Delta \theta_\mathrm{bias})$ and $h_E(\theta)$. Note that we estimate $\Delta \Phi_\mathrm{GW}$ from the dephasing between the $(2,2)$-modes, which is two times the difference between the phase trajectories of the \texttt{bhpwave} and corrupted models, $\Phi_B$ and $ \Phi_D$, respectively. The left panel includes results for the binary with masses $(M, \mu) = (10^6 M_\odot, 10 M_\odot)$, while the right panel is for a binary with $(M, \mu) = (5\times 10^6 M_\odot, 10 M_\odot)$. The dashed line just denotes a magnitude of 1. Thus, dephasings below one indicate subradian agreement between the two models, while ratios below one indicate that the systematic biases are below the intrinsic uncertainty of the measured parameters.

For both systems we see that the dephasing and biases in the intrinsic parameters tend to increase with spin. This is expected, since the error scales with the frequency, and higher spin systems will reach larger orbital frequencies and consequently possess larger errors. However, large dephasings do not necessarily guarantee that the systematic errors will significantly bias the measured parameters. For instance, in the $(M, \hat{a})=(10^6 M_\odot, 0.9)$ system, $\Delta \Phi_\mathrm{GW} \sim 10$ radians, yet the systematic biases remain smaller on the statistical parameter uncertainties.

Furthermore, for both systems, the biases in the extrinsic parameters are consistently below the parameter uncertainties. Again this is expected since the injected error only affects the intrinsic properties of the source. Interestingly, after accounting for the biases, we still generate small mismatches between the models across all of the tested systems. This indicates that our erroneous models, despite their large biases, can still capture most of the power in the EMRI gravitational wave signal (in the absence of noise). Thus, adiabatic waveforms may be sufficiently accurate for measuring the extrinsic parameters of EMRIs in the LISA data stream and systems with very low spins or retrograde orbits.

@@

We presented the theoretical and numerical methods behind \texttt{bhpwave}: a new Python-based adiabatic gravitational waveform generator for binary sources composed of a compact object undergoing a quasi-circular inspiral into a rotating massive black hole. To build this waveform model, we precomputed mass-independent trajectories for systems with initial separations $r_0/M \lesssim 60$ and Kerr spins $|a| \leq 0.9999$. Furthermore, we precomputed all potential waveform harmonic amplitudes for $\ell \leq 20$. By implementing the $E(3)$ boundary conditions in our numerical spline algorithm, we improved the precision of our interpolated flux data and trajectories by a couple orders of magnitude over traditional spline methods. We compute waveforms in both the time and frequency domains, and we find good agreement between these models. Furthermore, we compared our model against the gold-standard FEW waveform generator for $\hat{a} = e = 0$ and achieved mismatches $\mathcal{M} \sim 10^{-5}$.

In the future, we hope to potentially integrate our Kerr data in the FEW model in order to leverage its ability to run on GPUs.
